Mahotas: Open source software for scriptable computer vision

### 2012-08-30 paper.tex
TXT Add timings. Fix typos. Reorg

\documentclass{article}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{python}
\usepackage[%
    natbib=true,%
    backend=biber,%
    backref=true,%
    citestyle=authoryear-comp,%
    bibstyle=authoryear,%
    maxbibnames=24,%
    maxcitenames=2,%
]{biblatex}

@@

\section*{Abstract}
Mahotas is a computer vision library for Python. It contains traditional image
processing functionality such as filtering and morphological operations as well
as more modern computer vision functions for feature computation, including
local features.

@@

Mahotas is a computer vision library for the Python Programming Language
(versions 2.5 and up, including version 3). It operates on numpy
arrays~\citep{numpystructure}. Therefore, it uses all the infrastructure built
by that project for storing information. In particular, unlike libraries
written in the C~Language or in Java~\citep{Marcel:2010:TMP:1873951.1874254},
Mahotas does not need to define a data structure, but uses the numpy array
structure. Many basic manipulation functionality that would otherwise be part
of a computer vision library are handled by numpy, for example computing
averages and other simple statistics, handling multi-channel images, converting
between types (integer and floating point images are supported by mahotas,
whenever it is meaningful). For the user, this has the additional advantage
that they do not need to learn yet another set of functions.

On the other end, by integrating into the Python numeric ecosystem, users can
use other packages in a seamless way. In particular, mahotas does not implement
any machine learning functionality, but rather advises the user to use another,
specialised package, such as scikits-learn or milk.

@@

\begin{description}
\item[\textsc{surf}] Speeded-up Robust Features~\citep{eth_biwi_00517}. This
includes both keypoint detection and descriptor computation.
\item[features] Global feature descriptors. In particular, Haralick texture
features, Zernike moments, local binary patterns, and threshold adjacency
statistics (both the original~\citep{Hamilton2007} and the parameter-free
versions~\citep{Coelho2010}).
\item[Wavelet] Haar and Daubechies wavelets. Forward and inverse transforms.
\item[morphological functions] Erosion and dilation, as well as some more
complex operations built on these. There are both binary and grayscale
implementations of these operators.
\item[watershed] seeded watershed and distance map
transforms~\citep{felzenszwalb}.
\item[filtering] Gaussian filtering and general convolutions.
\item[polygon operations] convex hull, polygon drawing.
\end{description}

In general, functions operate on all data types (in some cases, the operation
is only meaningful for floating point arrays, for example). This is performed
without any extra memory copies. Mahotas is heavily optimised for both speed
and memory usage (it can be used with very large arrays).

@@

\begin{cplusplus}
PyObject* py_erode(PyObject* self, PyObject* args) {
    PyArrayObject* array;
    PyArrayObject* Bc;
    if (!PyArg_ParseTuple(args,"OO", &array, &Bc)) {
        return NULL;
    }
    PyArrayObject* res_a = (PyArrayObject*)PyArray_SimpleNew(
                                array->nd,
                                array->dimensions,
                                PyArray_TYPE(array));
    if (!res_a) return NULL;
    PyArray_FILLWBYTE(res_a, 0);
    switch(PyArray_TYPE(array)) {
#define HANDLE(type) \
    erode<type>(numpy::aligned_array<type>(res_a), \
                numpy::aligned_array<type>(array), \
                numpy::aligned_array<type>(Bc));

        HANDLE_INTEGER_TYPES();
#undef HANDLE
    ...
\end{cplusplus}

This functions retrieves the arguments, performs some sanity checks, performs a
bit of initialisation, and finally, switches in the input type with the help of
the \code{HANDLE\_INTEGER\_TYPES()} macro, which call the right specialisation
of the template that does the actual work. In this example \code{erode}
implements (binary) erosion:

\begin{cplusplus}
template<typename T>
void erode(numpy::aligned_array<T> res,
            numpy::aligned_array<T> array,
            numpy::aligned_array<T> Bc) {
    gil_release nogil;
    const unsigned N = res.size();
    typename numpy::aligned_array<T>::iterator iter = array.begin();
    filter_iterator<T> filter(res.raw_array(), Bc.raw_array());
    const unsigned N2 = filter.size();
    T* rpos = res.data();

    for (int i = 0;
                i != N;
                ++i, ++rpos, filter.iterate_both(iter)) {
        for (int j = 0; j != N2; ++j) {
            T arr_val = false;
            filter.retrieve(iter, j, arr_val);
            if (filter[j] && !arr_val) {
                goto skip_this_one;
            }
        }
        *rpos = true;
        skip_this_one: continue;
    }
}
\end{cplusplus}

The template machinery makes the functions that use it very simple and easy to
read. The only downside is that there is some expansion of code size when the
compiler instanciates the function for the several integer and floating point
types. Given the small size of these functions, this is not a big issue.

@@

The inner loop is as direct an implementation of erosion as one would wish for:
for each pixel in the image, look at its neighbours. If all are true, then set
the corresponding output pixel to \textbf{true} (else, skip it as it has been
initialised to zero).

@@

\textbf{Operating system}\\
Mahotas runs and is used on different versions of Unix (including Linux, SunOS,
and FreeBSD), Mac OS X, and Windows.

@@

Mahotas has been available in the Python Package Index since April~2010 and has
been downloaded over 20,000~times. This does not include any downloads from
other sources. Mahotas includes a full test suite (version 1.0 has 100\% test
coverage). There are no known bugs.

@@

>>>>

\documentclass{article}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{python}
\usepackage[%
    natbib=true,%
    backend=biber,%
    backref=true,%
    citestyle=authoryear-comp,%
    bibstyle=authoryear,%
    maxbibnames=24,%
    maxcitenames=2,%
]{biblatex}

@@

\section*{Abstract}
Mahotas is a computer vision library for Python. It contains traditional image
processing functionality such as filtering and morphological operations as well
as more modern computer vision functions for feature computation, including
interest point detection and local descriptors.

@@

Mahotas is a computer vision library for the Python Programming Language
(versions 2.5 and up, including version 3). It operates on numpy
arrays~\citep{numpystructure}. Therefore, it uses all the infrastructure built
by that project for storing information and performing basic manipulations and
computations. In particular, unlike libraries written in the C~Language or in
Java~\citep{Marcel:2010:TMP:1873951.1874254}, Mahotas does not need to define a
data structure, but uses the numpy array structure. Many basic manipulation
functionality that would otherwise be part of a computer vision library are
handled by numpy, for example computing averages and other simple statistics,
handling multi-channel images, converting between types (integer and floating
point images are supported by mahotas, whenever it is meaningful). For the
user, this has the additional advantage that they do not need to learn yet
another set of functions.

Additionally, by integrating into the Python numeric ecosystem, users can use
other packages in a seamless way. In particular, mahotas does not implement any
machine learning functionality, but rather advises the user to use another,
specialised package, such as scikits-learn or milk.

@@

\begin{description}
\item[\textsc{surf}] Speeded-up Robust Features~\citep{eth_biwi_00517}. This
includes both keypoint detection and descriptor computation.
\item[features] Global feature descriptors. In particular, Haralick texture
features, Zernike moments, local binary patterns, and threshold adjacency
statistics (both the original~\citep{Hamilton2007} and the parameter-free
versions~\citep{Coelho2010}).
\item[wavelet] Haar and Daubechies wavelets. Forward and inverse transforms.
\item[morphological functions] Erosion and dilation, as well as some more
complex operations built on these. There are both binary and grayscale
implementations of these operators.
\item[watershed] seeded watershed and distance map
transforms~\citep{felzenszwalb}.
\item[filtering] Gaussian filtering and general convolutions.
\item[polygon operations] convex hull, polygon drawing.
\end{description}

In general, functions operate on all data types. This is performed without any
extra memory copies. Mahotas is heavily optimised for both speed and memory
usage (it can be used with very large arrays).

@@

\begin{cplusplus}
PyObject* py_erode(PyObject* self, PyObject* args) {
    PyArrayObject* array;
    PyArrayObject* Bc;
    PyArrayObject* output;
    if (!PyArg_ParseTuple(args, "OOO", &array, &Bc, &output) ||
        !numpy::are_arrays(array, Bc, output) || !numpy::same_shape(array, output) ||
        !PyArray_EquivTypenums(PyArray_TYPE(array), PyArray_TYPE(Bc)) ||
        !PyArray_EquivTypenums(PyArray_TYPE(array), PyArray_TYPE(output)) ||
        PyArray_NDIM(array) != PyArray_NDIM(Bc)
    ) {
        PyErr_SetString(PyExc_RuntimeError, TypeErrorMsg);
        return NULL;
    }
    holdref r_o(output);

#define HANDLE(type) \
    erode<type>(numpy::aligned_array<type>(output), \
                numpy::aligned_array<type>(array), \
                numpy::aligned_array<type>(Bc));
    SAFE_SWITCH_ON_INTEGER_TYPES_OF(array, true);
#undef HANDLE
    ...
\end{cplusplus}

This functions retrieves the arguments, performs some sanity checks, performs a
bit of initialisation, and finally, switches in the input type with the help of
the \code{SAFE\_SWITCH\_ON\_INTEGER\_TYPES()} macro, which call the right
specialisation of the template that does the actual work. In this example
\code{erode} implements (binary) erosion:

\begin{cplusplus}
template<typename T>
void erode(numpy::aligned_array<T> res,
            numpy::aligned_array<T> array,
            numpy::aligned_array<T> Bc) {
    gil_release nogil;
    const int N = res.size();
    typename numpy::aligned_array<T>::iterator iter = array.begin();
    filter_iterator<T> filter(array.raw_array(), Bc.raw_array(), EXTEND_NEAREST, is_bool(T()));
    const int N2 = filter.size();
    T* rpos = res.data();

    for (int i = 0;
            i != N;
                ++i, ++rpos, filter.iterate_both(iter)) {
        T value = std::numeric_limits<T>::max();
        for (int j = 0; j != N2; ++j) {
            T arr_val = T();
            filter.retrieve(iter, j, arr_val);
            value = std::min<T>(value, erode_sub(arr_val, filter[j]));
        }
        *rpos = value;
    }
}
\end{cplusplus}

The template machinery makes the functions that use it very simple and easy to
read. The only downside is that there is some expansion of code size when the
compiler instantiates the function for the several integer and floating point
types. Given the small size of these functions, this is not a big issue.

@@

The inner loop is as direct an implementation of erosion as one would wish for:
for each pixel in the image, look at its neighbours, subtract the filter value,
and compute the minimum of this operation.

\subsection{Efficiency}

\begin{table}
\centering
\begin{tabular}{lrr}
\toprule
Operation & mahotas & pymorphy\\
\midrule
erode       &     9.90 &    13.06 \\
dilate      &    10.48 &     8.17 \\
open        &    20.26 &    21.05 \\
cwatershed  &   181.47 & 50211.56 \\
center mass &     6.39 &       NA \\
daubechies  &    16.92 &       NA \\
haralick    &   276.90 &       NA \\
\bottomrule
\end{tabular}
\caption{Efficiency Results for mahotas and pymorph. Shown are values as
multiples of the time that \code{numpy.max(f)} takes.}
\label{tab:efficiency}
\end{table}

Table~\ref{tab:efficiency} shows timings for different operations. These were
normalized to multiples of the time it takes to go over the image and find its
maximum pixel value (this was done using \code{numpy.max(image}). The
measurements shown were obtained on an Intel 64~bit system, running Ubuntu
Linux. However, due to the normalization, measurements obtained on another
system (Intel 32~bits running Mac OS) were qualitatively similar (data not
shown).

The comparison is against Pymorph~\citep{DougLotu:03}, which is a pure Python
implementation of some of the same functions. We can see that simple
morphological operations (erosion, dilation, opening) take a similar amount of
time in mahotas and pymorph (pymorph intelligently uses numpy operations for
these). However, some more complex operations such as watershed are much faster
in mahotas.

\bigskip

@@

\textbf{Operating system}\\
Mahotas runs and is used on different versions of Unix (including Linux, SunOS,
and FreeBSD), Mac OS X, and Windows.\footnote{Christoph Gohlke has been
instrumental in providing Windows packages as well as several fixes for that
platform.}

@@

@@

Python is an excellent language for scientific programming because of the
inherent properties of the language and because of the infrastructure that has
been built around the numpy project. Mahotas works in this environment to
provide the user with image analysis and computer vision functionality.

Mahotas has been available in the Python Package Index since April~2010 and has
been downloaded over 20,000~times. This does not include any downloads from
other sources. Mahotas includes a full test suite. There are no known bugs.
